---
title: "Economics of Underdevelopment Homework – Egypt"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Setup code chunk: load libraries and set global options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Install packages if not already installed 
# install.packages("WDI")
# install.packages("ggplot2")
# install.packages("reshape2")
# install.packages("dplyr")

```{r setup, include=FALSE}
library(WDI)
library(ggplot2)
library(reshape2)
library(dplyr)
WDIsearch("GDP per capita")
```

# 1. GDP per Capita

# -- 1.1 GDP per Capita (Chain-Linked Volume) for Egypt --

```{r}

# Set variables
egypt_code <- "EGY"  # ISO-3 code for Egypt
gdp_chain_indicator <- "NY.GDP.PCAP.KD"  # GDP per capita, constant US$

# Download data from 1993 to 2023 (about 30 years)
egypt_gdp <- WDI(
  country  = egypt_code, 
  indicator = gdp_chain_indicator, 
  start    = 1993, 
  end      = 2023
)

# Quick check
head(egypt_gdp)

# Plot GDP per capita over time
ggplot(egypt_gdp, aes(x = year, y = NY.GDP.PCAP.KD)) +
  geom_line(color = "blue", linewidth = 2) +
  labs(
    title = "Egypt GDP per Capita (Chain-Linked Volume, 1993-2023)",
    x = "Year",
    y = "GDP per Capita (constant US$)"
  ) +
  theme_minimal()

```

# -- 1.2 GDP per Capita (PPP) Comparison in 2019 --

```{r}
# Define neighbors. You can choose whichever neighbors are most relevant.
# Common neighbors for Egypt might be: Israel (ISR), Libya (LBY), Sudan (SDN), Saudi Arabia (SAU), Jordan (JOR)
neighbors <- c("LBY", "ISR", "SAU", "SDN", "JOR")

gdp_ppp_indicator <- "NY.GDP.PCAP.PP.CD"  # GDP per capita, PPP (current international $)

# Download data for Egypt + neighbors in 2019
countries_ppp <- c(egypt_code, neighbors)
gdp_ppp_2019 <- WDI(
  country   = countries_ppp, 
  indicator = gdp_ppp_indicator, 
  start     = 2019, 
  end       = 2019
)

# Check the data
gdp_ppp_2019

# Plot a comparison bar chart
ggplot(gdp_ppp_2019, aes(x = country, y = NY.GDP.PCAP.PP.CD, fill = country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "GDP per Capita (PPP) in 2019: Egypt vs. Neighbors",
    x = "Country",
    y = "GDP per Capita (PPP, current int'l $)",
    fill = "Country"
  ) +
  theme_minimal()
```

# 2.Population

# -- 2.1 Total Population --
```{r}
# Download total population data for Egypt from 1960 to the latest available
pop_data <- WDI(
  country   = egypt_code,
  indicator = "SP.POP.TOTL",
  start     = 1960,
  end       = 2023
)
head(pop_data)

# Plot total population over time
ggplot(pop_data, aes(x = year, y = SP.POP.TOTL)) +
  geom_line(color = "red", linewidth = 2) +
  labs(
    title = "Egypt Total Population (1960-2023)",
    x = "Year",
    y = "Total Population"
  ) +
  theme_minimal()
```

# -- 2.2 Age Structure --

```{r}
# Indicators for age groups:
# SP.POP.0014.TO.ZS = Ages 0-14 (% of total)
# SP.POP.1564.TO.ZS = Ages 15-64 (% of total)
# SP.POP.65UP.TO.ZS = Ages 65+ (% of total)
age_structure_data <- WDI(
  country = egypt_code,
  indicator = c("SP.POP.0014.TO.ZS", "SP.POP.1564.TO.ZS", "SP.POP.65UP.TO.ZS"),
  start = 1960,
  end   = 2023
)
head(age_structure_data)
# Subset just the columns we need (year + the 3 age-share columns)
age_data <- age_structure_data[, c("year", "SP.POP.0014.TO.ZS", "SP.POP.1564.TO.ZS", "SP.POP.65UP.TO.ZS")]
# Melt the data for plotting
age_melt <- melt(age_data, id.vars = "year", variable.name = "age_group", value.name = "percentage")
# Rename the age groups for clarity (optional)
age_melt$age_group <- factor(age_melt$age_group,
                             levels = c("SP.POP.0014.TO.ZS", "SP.POP.1564.TO.ZS", "SP.POP.65UP.TO.ZS"),
                             labels = c("0-14", "15-64", "65+")
)

# Plot the age structure over time (stacked area)
ggplot(age_melt, aes(x = year, y = percentage, fill = age_group)) +
  geom_area(alpha = 0.7) +
  labs(
    title = "Egypt Age Structure Over Time",
    x = "Year",
    y = "Percentage of Population",
    fill = "Age Group"
  ) +
  theme_minimal()
```

# 3. Production Structure

# -- 3.1 Employment Shares --

```{r}
# Indicators:
# SL.AGR.EMPL.ZS = Employment in agriculture (% of total)
# SL.IND.EMPL.ZS = Employment in industry (% of total)
# SL.SRV.EMPL.ZS = Employment in services (% of total)
emp_data <- WDI(
  country = egypt_code,
  indicator = c("SL.AGR.EMPL.ZS", "SL.IND.EMPL.ZS", "SL.SRV.EMPL.ZS"),
  start = 1990,
  end   = 2023
)
head(emp_data)

# Plot employment shares
ggplot(emp_data, aes(x = year)) +
  geom_line(aes(y = SL.AGR.EMPL.ZS, color = "Agriculture"), linewidth = 2) +
  geom_line(aes(y = SL.IND.EMPL.ZS, color = "Industry"), linewidth = 2) +
  geom_line(aes(y = SL.SRV.EMPL.ZS, color = "Services"), linewidth = 2) +
  labs(
    title = "Egypt Employment Share by Sector (1990-2023)",
    x = "Year",
    y = "Percentage of Total Employment",
    color = "Sector"
  ) +
  theme_minimal()
```

# -- 3.2 Value Added Shares --

```{r}
# Indicators:
# NV.AGR.TOTL.ZS = Agriculture value added (% of GDP)
# NV.IND.TOTL.ZS = Industry value added (% of GDP)
# NV.SRV.TOTL.ZS = Services value added (% of GDP)
val_added_data <- WDI(
  country   = egypt_code,
  indicator = c("NV.AGR.TOTL.ZS", "NV.IND.TOTL.ZS", "NV.SRV.TOTL.ZS"),
  start     = 1990,
  end       = 2023
)
head(val_added_data)

# Plot value added shares
ggplot(val_added_data, aes(x = year)) +
  geom_line(aes(y = NV.AGR.TOTL.ZS, color = "Agriculture"), linewidth = 2) +
  geom_line(aes(y = NV.IND.TOTL.ZS, color = "Industry"), linewidth = 2) +
  geom_line(aes(y = NV.SRV.TOTL.ZS, color = "Services"), linewidth = 2) +
  labs(
    title = "Egypt Value Added Share by Sector (1990-2023)",
    x = "Year",
    y = "Percentage of GDP",
    color = "Sector"
  ) +
  theme_minimal()
```

# 4. Open Economy

```{r}
WDIsearch("current account")
WDIsearch("trade balance")
WDIsearch("net foreign assets")
```

# 1. Current Account Balance (% of GDP)
# 2. Trade Balance (% of GDP) – calculated from exports and imports data
# 3. Net Foreign Asset Position (% of GDP) – calculated from net foreign assets and GDP data
# Set the analysis period (last 30 years)

```{r}
# Set the analysis period (last 30 years)
start_year <- 1960
end_year   <- 2023
```

# 1. Current Account (% GDP)
# Alternative Indicator: BN.CAB.XOKA.GD.ZS - Current account balance (% of GDP)
```{r}
current_account <- WDI(country = "EGY", 
                       indicator = "BN.CAB.XOKA.GD.ZS", 
                       start = start_year, 
                       end = end_year)
# Check the first few rows to verify data
head(current_account)
```

# 2. Trade Balance (% GDP)
# Since there is no direct trade balance (% GDP) indicator,
# we calculate it using exports, imports, and GDP data.
# Exports: NE.EXP.GNFS.CD - Exports of goods and services (current US$)
# Imports: NE.IMP.GNFS.CD - Imports of goods and services (current US$)
# GDP: NY.GDP.MKTP.CD - GDP (current US$)
# Download exports data for Egypt
```{r}
exports <- WDI(country = "EGY", 
               indicator = "NE.EXP.GNFS.CD", 
               start = start_year, 
               end = end_year)
# Download imports data for Egypt
imports <- WDI(country = "EGY", 
               indicator = "NE.IMP.GNFS.CD", 
               start = start_year, 
               end = end_year)
# Download GDP data (current US$) for Egypt
gdp_current <- WDI(country = "EGY", 
                   indicator = "NY.GDP.MKTP.CD", 
                   start = start_year, 
                   end = end_year)
# Merge exports, imports, and GDP data by country code and year
trade_data <- exports %>%
  inner_join(imports, by = c("iso2c", "country", "year")) %>%
  inner_join(gdp_current, by = c("iso2c", "country", "year"))
# Calculate Trade Balance as (% of GDP)
# Formula: ((Exports - Imports) / GDP) * 100
trade_data <- trade_data %>%
  mutate(trade_balance_pct = (NE.EXP.GNFS.CD - NE.IMP.GNFS.CD) / NY.GDP.MKTP.CD * 100)
# Check calculated trade balance
head(trade_data[, c("iso2c", "country", "year", "trade_balance_pct")])
```

# 3. Net Foreign Asset Position (% GDP)
# There is no direct indicator in % of GDP for net foreign assets.
# We use the indicator for net foreign assets in current US$
# and calculate the ratio relative to GDP (current US$).
# Indicator for Net Foreign Assets (current US$):
# NW.NFA.TO - Net foreign assets (current US$)
```{r}
nfa <- WDI(country = "EGY", 
           indicator = "NW.NFA.TO", 
           start = start_year, 
           end = end_year)
# Merge net foreign assets with GDP data (both in current US$)
nfa_data <- nfa %>%
  inner_join(gdp_current, by = c("iso2c", "country", "year"))
# Calculate Net Foreign Asset Position as (% of GDP)
# Formula: (Net Foreign Assets / GDP) * 100
nfa_data <- nfa_data %>%
  mutate(nfa_pct = (NW.NFA.TO / NY.GDP.MKTP.CD) * 100)
# Check calculated net foreign asset position
head(nfa_data[, c("iso2c", "country", "year", "nfa_pct")])
```

# Combine all three indicators into one data frame
# Merge the datasets by year for easy comparison.
```{r}
combined <- current_account %>%
  select(iso2c, country, year, current_account_pct = BN.CAB.XOKA.GD.ZS) %>%
  inner_join(trade_data %>% select(iso2c, country, year, trade_balance_pct), 
             by = c("iso2c", "country", "year")) %>%
  inner_join(nfa_data %>% select(iso2c, country, year, nfa_pct), 
             by = c("iso2c", "country", "year"))
# View combined data
head(combined)
```

# Plotting the Indicators Over Time
```{r}
# Plot 1: Current Account (% GDP)
ggplot(combined, aes(x = year, y = current_account_pct)) +
  geom_line(color = "blue", linewidth = 2) +
  geom_point(color = "blue") +
  labs(title = "Egypt Current Account (% GDP)",
       x = "Year",
       y = "Current Account (% GDP)") +
  theme_minimal()
# Plot 2: Trade Balance (% GDP)
ggplot(combined, aes(x = year, y = trade_balance_pct)) +
  geom_line(color = "red", linewidth = 2) +
  geom_point(color = "red") +
  labs(title = "Egypt Trade Balance (% GDP)",
       x = "Year",
       y = "Trade Balance (% GDP)") +
  theme_minimal()
# Plot 3: Net Foreign Asset Position (% GDP)
ggplot(combined, aes(x = year, y = nfa_pct)) +
  geom_line(color = "green", linewidth = 2) +
  geom_point(color = "green") +
  labs(title = "Egypt Net Foreign Asset Position (% GDP)",
       x = "Year",
       y = "Net Foreign Assets (% GDP)") +
  theme_minimal()
```
